---
title: "Introduction to R"
---

## R workspace

In R, one works in an area called the *workspace*.
The workspace is a working environment where objects are created and manipulated.
Objects commonly kept in the workspace are __(a)__ entire data sets (i.e. tibbles)
and __(b)__ the output of statistical analyses.
It is also relatively common to keep programs (i.e., functions) that do
special project-related tasks within the workspace.
Use `ls()` to display the names of the objects in the workspace.

Within the workspace, one removes objects x, y, z using the `rm()` function:
`rm(x, y, z)` if you fail to save the workspace after adding or modifying objects you create
in the current session, they will NOT be there next time you start R and load the specific workspace. 

> For reproducibility, it is recomended not to save workspace when quiting R.  

To save current workspace, use the RStudio "Save Workspace As" option under save icon from the Environment pane to specify where to save the workspace.

To get more information on any specific named function, for example solve, the command is either `help()` or you can prepend command of interest with question mark `?`.

```{r getting-help}
help(solve)
?solve()
```

Searches of help files can be conducted using the `help.search()` function.
For instance, to find functions related to regression one would type:

```{r help-search}
help.search("regression")
```

##Run command
Use `Cntr/Cmd + Enter` to evaluate the code or click the Run button:
```{r evaluate-code}
2+2 # Cntr/Cmd + Enter
```

##R functions

Entering command with its arguments returns the sum of all the values present in its arguments.
```{r functions}
x <- c(1.1, 2, pi) # vector of arguments
sum(x) # calculates sum of vector x
```

You can look at the function arguments using help.
```{r}
?sum()
```

`sum()` function has two arguments: `...` __dot-dot-dot a.k.a ellipsis__ and default argument value - `na.rm = FALSE`, you can change it!

- A collection of similar functions is a package
- you can download many thousands of packages
- from repositories (CRAN, bioconductor): `install.packages("packagename")`
- takes directly from CRAN repository
- tidyverse is a package that is itself a
- collection of packages that work well together
- in order to use the functions in a package
- you have to first load the package: library(packagename)
- each fun has a help page, most packages have tutorials and such.

Next code helps you to install all your packages to a new machine or R version.
this is very useful.

In the old version/machine run:
```{r eval=FALSE}
installed <- as.data.frame(installed.packages())
write.csv(installed, "installed_previously.csv")
```

And in the new one:
```{r}
installedPreviously <- read.csv("installed_previously.csv")
baseR <- as.data.frame(installed.packages())
toInstall <- setdiff(installedPreviously$Package, baseR$Package)
install.packages(toInstall) 
```


## Names
We can assign values to names, thus creating objects.
`<-` does the assigning, shortcut: Alt + - (the minus sign). 
All statements where you create objects have the same form: `object_name <- value`
When reading that code say "object name gets value".

- Object names must start with a letter, and
- can only contain letters, numbers, _, and .
- no spaces
- case-sensitive
- a long name: this_is_a_long_name_01
- use TAB to autocomplete

```{r}
apples <- 2 #two apples
oranges <- 3 #three oranges
inventory <- apples + oranges
inventory #inspect the object
```


or you can use the function sum()
```{r}
inventory <- sum(apples, oranges)
```

Here we changed the name (or, more accurately, created a new name for the same object, thus cloning the object):

```{r}
apelsinid <- oranges  
```

We can also change the object behind a name.
The last change stands.

```{r}
apples <- apples + 4
inventory <- apples + oranges
inventory
```


<!-- ### Fun -->
<!-- As we get more apples and oranges, -->
<!-- - we want a way to reduce typing. -->
<!-- - for this we write functions, -->
<!-- - which hide the code from the user -->

<!-- ```{r} -->
<!-- inventory <- function(apples, oranges) { -->
<!--   inventory <- apples + oranges -->
<!--   inventory -->
<!-- } -->
<!-- ``` -->

<!-- Run this code and see how the new fun appears in the Global Enviroment.  -->
<!-- Let's use our new function -->
<!-- ```{r} -->
<!-- inventory(3, 4) -->
<!-- ``` -->

<!-- or equivalently -->

<!-- ```{r} -->
<!-- inventory(apples = 3, oranges = 4) -->
<!-- ``` -->



<!-- OK, write a function that calculates standard error -->
<!-- from standard deviation (sd) -->
<!-- sd() is a native R fun for standard deviation - -->
<!-- look it up in help! it has 2 arguments. -->
<!-- ```{r} -->
<!-- SEM=SD/sqrt(N) -->
<!-- N - length(x) -->

<!-- SEM0<- function(x) { -->
<!--   SD<- sd(x)  -->
<!--   SEM<- SD/sqrt(length(x)) -->
<!--   SEM -->
<!-- } -->
<!-- ``` -->

<!-- x is just a placeholder for any object we want to insert -->
<!-- so lets input a vector of numbers to SEM() -->
<!-- ```{r} -->
<!-- numbers<- c(2, 3.4, 54, NA, 3) #c() means "combine", NA means "any number" or "not available" -->

<!-- SEM0(numbers) -->

<!-- SEM01<- function(x) { -->
<!--   SD<- sd(x, na.rm = TRUE)  -->
<!--   SEM<- SD/sqrt(length(x)) -->
<!--   SEM -->
<!-- } -->
<!-- SEM01(numbers) -->
<!-- length(numbers) -->
<!-- ``` -->


<!-- While SD removes NA-s, the length() does not. This leads to errors! -->
<!-- ```{r} -->
<!-- SEM1<- function(x) { -->
<!--   x<- na.omit(x) #you want NAs to be removed -->
<!--   SD<- sd(x)  -->
<!--   SEM<- SD/sqrt(length(x)) -->
<!--   SEM -->
<!-- } -->

<!-- SEM1(numbers) #this works better! -->

<!-- SEM1<- function(x) { -->
<!--   SD<- sd(x, na.rm = TRUE) # na.rm is an argument of the sd() fun -->
<!--   SEM<- SD/sqrt(length(na.omit(x))) #we nest 2 functions (length() and na.omit()) inside our function -->
<!--   SEM -->
<!-- } -->
<!-- SEM1(numbers) -->

<!-- SEM2<-function(x) { -->
<!--   SD<- sd(na.omit(x)) -->
<!--   SEM<- SD/sqrt(length(na.omit(x) ) ) -->
<!--   SEM -->
<!-- } -->
<!-- SEM2(numbers) -->
<!-- ``` -->



<!-- Harjutus: sd ilma sd() kasutamata -->
<!-- ```{r} -->
<!-- SEM3 <- function(x) { -->
<!--   x <- na.omit(x) -->
<!--   N <- length(x) -->
<!--   SD <- sqrt(sum( (x-mean(x) )**2) / ( N-1 ) ) -->
<!--   SEM <- SD/sqrt(N) -->
<!--   SEM -->
<!--   } -->

<!-- SD3 <- function(x) { -->
<!--   x <- na.omit(x) -->
<!--   N <- length(x) -->
<!--   SD <- sqrt(sum( (x-mean(x) )**2) / ( N-1 ) ) -->
<!--   SD -->
<!-- } -->
<!-- ``` -->


<!-- ```{r} -->
<!-- a <- rnorm(10, 10, 3) -->
<!-- SD3(a) -->
<!-- sd(a) -->
<!-- SEM3(a) -->
<!-- SEM2(a) -->
<!-- SEM1(a) -->
<!-- ``` -->


<!-- ### Lets start with installing a package -->
<!-- ```{r} -->
<!-- install.packages("tidyverse") -->
<!-- ``` -->

<!-- and load it -->
<!-- ```{r} -->
<!-- library(tidyverse) -->
<!-- ``` -->



<!-- #now its time to read in a data table (csv) -->
<!-- #NB! it is good practice to put all your library()  -->
<!-- #calls at the beginning of your script (if you share scripts) -->
<!-- library(gapminder) -->
<!-- #lets look at it in the Packages panel -->
<!-- # gapminder is the "gapminder data" -->
<!-- gap1<- gapminder -->

<!-- write_csv(gap1, "data/gapminder1.csv") -->
<!-- #see ?write_csv -->

<!-- gap<-read_csv("data/gapminder1.csv") -->
<!-- gap -->

<!-- > Mis (fuck) see on? -->

<!-- #install gotta read em all as R studio addin -->
<!-- install.packages("devtools") -->
<!-- devtools::install_github("Stan125/GREA") -->

<!-- when reading in csv files saved from Excel use read_csv2() or read.csv2() and -->
<!-- for writing in Excel friendly format (columns separated by ;) use write.csv2() -->

<!-- When reading in a file, always check for type of data in the columns, -->
<!-- on the NAs, on the unique values. -->

<!-- library(VIM) # võib olla ajaraisk! -->
<!-- diabetes <- read.table(file = "data/diabetes.csv", sep = ";", dec = ",", header = TRUE) -->
<!-- str(diabetes) -->
<!-- aggr(diabetes, prop=FALSE, numbers=T) -->

<!-- how many non-NA-containing rows we actually have (#377) -->

<!--   nrows <- nrow(diabetes) -->
<!--   ncomplete <- sum(complete.cases(diabetes)) -->
<!--   ncomplete #136 -->
<!--   ncomplete/nrows #34% -->

<!--   #How many NAs in each Var? -->
<!--  sapply(diabetes, function(x) sum(is.na(x)))  -->

<!-- set arbitrary rules and check how many cells, rows, or columns pass and fail each rule -->

<!--  library(validate) -->
<!--  diabetes %>% check_that( -->
<!--    ratio > 0.5,  -->
<!--    mean(hdl, na.rm=TRUE) > 0, -->
<!--    chol > 180 | glyhb < 8  -->
<!--  ) %>% summary() -->

<!-- plot the NA-s side-by-side for the variables in the diabetes dataset -->
<!--   VIM::matrixplot(diabetes)  -->

<!--   plot the NAs of 2 Vars -->
<!--   VIM::marginplot(diabetes[c("glyhb","bp.2s")],  -->
<!--                   pch=c(20), alpha=0.5,  -->
<!--                   col=c("darkgray", "red", "blue")) -->


<!-- ### Small tables can be created in R -->
<!-- #row-wise -->
<!-- df <- tribble( -->
<!--   ~colA, ~colB, -->
<!--   "a",   1, -->
<!--   "b",   2, -->
<!--   "c",   3 -->
<!-- ) -->
<!-- #or column-wise  -->
<!-- df1 <- tibble(colA = c("a", "b", "c"),  -->
<!--                   colB = c(1, 2, 3) ) -->
<!-- df == df1 -->
<!-- all.equal(df, df1) -->
<!-- add_row(df, colA = 4, colB = 0, .before = 3) #adds row to a tibble -->
<!-- add_row(df1, colA = 4, colB = 0)  -->

<!-- #bind 2 tables by rows or by columns: two sets of functions -->
<!-- df1.1 <- tibble(colA = "d",  -->
<!--                 colB =  4) -->

<!-- df1.2 <- bind_rows(df1, df1.1) -->
<!-- df1.3 <- rbind(df1, df1.1) -->
<!-- all.equal(df1.2, df1.3) #TRUE -->

<!-- df1.4 <- bind_cols(df1, df1.1)#error -->
<!-- df1.5 <- cbind(df1, df1.1) #guesses your true intention --- dangerous -->


<!-- df2 <- tribble( -->
<!--   ~ColC, ~ColD, -->
<!--   "d",   4 -->
<!-- ) -->

<!-- df4 <- bind_cols(df, df2) #error -->
<!-- df4 <- cbind(df, df2) #works by guessing your true intention -->
<!-- df4.2 <- bind_rows(df, df2) #works by guessing your true intention -->
<!-- df4.3 <- rbind(df, df2) #error -->

<!-- df2.1 <- tribble( -->
<!--   ~ColC, ~ColD, -->
<!--   "d",   4, -->
<!--   "e",  5, -->
<!--   "f",   6) -->
<!-- df4.1 <- bind_cols(df, df2.1) #works -->


<!-- #########tibble and data frame are R-s table formats -->
<!-- class (df) #[1] "tbl_df"     "tbl"        "data.frame" -->
<!-- #tibble is the newer form -->

<!-- df <- as.data.frame(df) #converts to data frame -->
<!-- class(df)# [1] "data.frame" -->
<!-- #the older form - some older functions need it -->
<!-- df<- as_tibble(df) #back to tibble -->


<!-- a <- matrix(1:20, ncol=4) -->
<!-- df_cl <- as.data.frame(a) -->
<!-- df_ti <- as_tibble(df_cl) -->
<!-- df_dp <- as_data_frame(a) -->
<!-- all.equal(df_ti, df_dp) -->
<!-- is.tibble(df_dp) -->

<!-- #########################################################types of objects -->
<!-- R operates on named data structures. The simplest such structure is the numeric -->
<!-- vector, which is a single entity consisting of an ordered collection of numbers. -->

<!-- types of r objects: vector, list, data frame, matrix -->
<!-- vector: a directional collection of numbers, characters, etc: v <- c(1, 3, "s", NA) -->
<!-- df: a list of equal length vectors -->
<!-- matrix: a list of equal length numeric vectors -->
<!-- list: a collection of r objects (including other lists --- lists within lists). -->


<!-- types of data inside r objects: numeric, integer, logical, character, factor (ordered and unordered) -->
<!-- coercing between data types: -->
<!-- matrix() -->
<!-- as.matrix() -->
<!-- as.vector() -->
<!-- as.factor() -->
<!-- as.character() -->

<!-- types of missing data: NA - unknown value; NaN - not a number (no value is possible) -->

<!-- NA + 3 = NA -->
<!-- ####################################### -->
<!-- indexing --- extracting parts of tables, vectors, lists -->

<!-- vector indexing in one dimension: counts positions inside the vector -->
<!-- my_vector <- 2:5 -->
<!-- my_vector[c(1,3)] -->
<!-- my_vector[-1] -->
<!-- my_vector[c(-1, -3)] -->
<!-- my_vector[3:5] -->

<!-- my_vector + 2 #each element of the vector is used, one after other.  -->
<!-- #2 is a vector of length one, it is recycled for the length of the longer vector. -->
<!-- my_vector2 <- c(1, 2) -->
<!-- my_vector + my_vector2 #2+1=3; 3+2=5; 4+1=5, 5+2=7 --- recycling again! -->
<!-- #no warning -->

<!-- my_vector3 <- c(1,2,3) -->
<!-- my_vector + my_vector3 #3 5 7 6  -->
<!-- #warning: longer object length is not a multiple of shorter object length -->

<!-- #data frame indexing in two dimensions: rows and columns -->
<!-- df -->
<!-- df[1,] -->
<!-- df[,2] #tibble -->
<!-- df$colB #vector -->
<!-- df1 <- as.data.frame(df) -->
<!-- df1[,2] #vector -->
<!-- df1[2,2] #vector -->
<!-- df1[2, c(1,2)] -->
<!-- df[2,2]#tibble -->
<!-- df$colB[df$colA != "a"] #a filter of colB by colA value -->
<!-- df$colA[df$colB > 1] -->

<!-- #list indexing in "three" dimensions: [ ] and [[ ]]. -->
<!-- my_list <- list(a=tibble(colA=c("A", "B"), colB=c(1,2)), b=c(1, NA, "s")) -->
<!-- #this list has two elements, a df called "a" and a character vector called "b". -->
<!-- str(my_list) -->

<!-- my_list[[1]] #class is df  --- we extracted a df from the list -->
<!-- my_list$a #class is df --- $ does the same thing as [[ ]] -->
<!-- my_list[1] #class is list (of one element ---  -->
<!-- #we took out all elements but the first, but the result is still list) -->

<!-- my_list$a[2,] #class is df -->

<!-- my_list[[1]][1,] #the first list object extracted,  -->
<!-- #and then the first row of that is extracted --- class: df -->

<!-- For lists, one generally uses [[ to select any single element, -->
<!-- whereas [ returns a list of the selected elements. -->
<!-- The [[ form allows only a single element to be selected using integer or -->
<!-- character indices, whereas [ allows indexing by vectors. -->
<!-- Note though that for a list, the index can be a vector and each element -->
<!-- of the vector is applied in turn to the list, the selected component, the -->
<!-- selected component of that component, and so on. -->
<!-- The result is still a single element. -->


