---
title: "Regular expression"
  output: html_document
---


A pattern that we use to search in strings can be strict (like a nucleotide string "ACAGC") or "fuzzy" ("ACNGY" where N is any nucleotide and Y is C or T). To create fuzzy search patterns we use a system called **regular expression**. As regular expression is both hugely flexible and insanely complicated, we can only scratch the surface here.

A [regular expression](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html) is a pattern that describes a set of strings. [Regular Expressions as used in R](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html)

+ Most characters, including all letters and digits, are regular expressions that match themselves. 
+ `.` matches any single character.
+ You can refer also to a character class, which is a list of characters enclosed between `[` and `]`, e.g. `[[:alnum:]]` is same as `[A-z0-9]`.
+ Most common character classes:
      - `[:alnum:]` includes alphanumerics (`[:alpha:]` and `[:digit:]`); 
      - `[:alpha:]`, includes alphabetic characters (`[:upper:]` and `[:lower:]` case);
      - `[:punct:]` includes punctuation characters ! " # $ % & ' ( ) * + , - . / : ; < = > ? @ [ \ ] ^ _ ` ` { | } ~.; 
      - `[:blank:]` includes space and tab; etc. 
+ The metacharacters in regular expressions are `. \ | ( ) [ { ^ $ * + ?`, whether these have a special meaning depends on the context. When matching a metacharacter as a regular character, precede it with a double backslash `\\`.
+ Repetition quantifiers put after regex specify how many times regex is matched: `?`, optional, at most once; `*`, zero or more times; `+`, one or more times; `{n}`, n times; `{n,}`, n or more times; `{n,m}`, n to m times. 
+ ^ anchors the regular expression to the start of the string.
+ $ anchors the the regular expression to end of the string.

## Common operations with regular expressions

+ Locate a pattern match (positions)
+ Extract a matched pattern
+ Identify a match to a pattern
+ Replace a matched pattern

**Get string length**
the number of characters in the string:
```{r}
x <- c("reformation", "post", "TLC")
nchar(x)
```


> Tidyverse's `stringr` library  produces outputs than can easily be used as inputs.

 `str_locate` and `str_extract`:
```{r}
library(stringr)
y <- c("AreaShape_Area", "AreaShape_Perimeter", "AreaShape_MajorAxisLength", "Intensity_MADIntensity")
str_locate(y, "[[:alnum:]]*$") # if we want only start and end positions
```
Note that, `str_locate` returns more conveniently a matrix (instead of a list as in regexpr).

```{r}
str_extract(y, "[[:alnum:]]*$") # if we want actual strings
```
Seems nice and tidy.


`str_locate_all` and `str_extract_all`:
```{r}
str_locate_all(y, "e") # to locate all start positions and lengths of letter "e"
```

```{r}
str_extract_all(y, "[[:upper:]]") # to extract all upper case letters
```


## Find matching elements in vector

`str_subset()` from `stringr` library:
```{r}
str_subset(y, "Area") # note that strings and pattern have changed positions
```

 `str_detect()`:
```{r}
str_detect(y, "Area")
```

##Find and replace

```{r, eval=FALSE}
x<- c("apple", "ananas", "banana")

#replaces all a-s at the beginning of strings with e-s
str_replace(x, "^a", "e") 

# str_replace only replaces at the first occurence at each string
str_replace(x, "a", "e") 

#str_replace_all replaces all a-s anywhere in the strings
str_replace_all(x, "a", "e") 

#replaces a and the following character at the end of string with nothing (i.e. deletes 2 chars)
str_replace(x, "a.$", "")

#replaces a-s or s-s at the end of string with e-s
str_replace(x, "(a|s)$", "e")

#replaces a-s or s-s anywhere in the string with e-s
str_replace_all(x, "a|s", "e")

#remove all numbers. 
y<-c("as1", "2we3w", "3e")
str_replace_all(y, "\\d", "") 

#remove everything, except numbers. 
str_replace_all(y, "[A-Za-z_]", "") 
```


To force to only match a complete string:

```{r}
x<- c("apple", "apple pie")
str_view(x, "^apple$")
```

**patterns that match more than one character:**


```{r, eval=FALSE}

. (dot): any character apart from a newline.

\\d: any digit.

\\s: any whitespace (space, tab, newline).

\[abc]: match a, b, or c.

\[!abc]: match anything except a, b, or c.

To create a regular expression containing \d or \s, you???ll need to escape the \ for the string, so you???ll type "\\\\d" or "\\\\s".

abc|d..f will match either "abc", or "deaf". 
```


```{r}
str_view(c("abc", "xyz"), "abc | xyz")
```

if precedence gets confusing, use parentheses to make it clear what you want:

```{r}
str_detect(c("grey", "gray"), "gr(e|a)y")
str_detect(c("grey", "gray"), "gr(?:e|a)y") #unequivocal vers. of the above

```

