---
output: html_document
---

## Tidyverse

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
library(tidyverse)
library(ggthemes)
library(readr)
```

### Tidy data

To standardize data analysis, you must start by standardizing data structure.
Tidy data arranges values so that the relationships between variables in a data set will parallel the relationship between vectors in R???s storage objects. R stores tabular data as a data frame, a list of atomic vectors arranged to look like a table. Each column in the table is a vector. 

In tidy data, each variable in the data set is assigned to its own column, i.e., its own vector in the data frame. As a result, you can extract all the values of a variable  by extracting the column vector that contains the variable, i.e. table1$cases. Because R does vector calculations element by element, it is fastest when you compare vectors directly side-by-side.

-	value is the result of a single measurement (167 cm). = cell
-	variable is what you measure (length, height), or a factor (sex, treatment).  = column
-	observation or data point is a set of measurements that made under similar conditions (John???s height and weight measured on 23.04.2012). = row
-	Observational unit (who or what was measured): subject no. 1, etc. = 1st column
-	Type of observational unit: humans, mice, cell lysates, etc. = table

> Tidy data: each value is in its own ???cell???, each variable in its own column, each observation in its own row, and each type of observational unit in its own table - useful for grouping, summarizing, filtering, and plotting. In a tidy table the order of columns is: 1. Observational unit 2. Factors & everything that was not measured (values fixed at experimental planning stage) 3. Measured Vars. 

Keeping the data in this form allows multiple tools to be used in sequence. 

NB! There are always more possible Vars in your data than were measured ??? do weight and height and get BMI as a bonus.

### Melt data into the long format

reshape2::melt(df) - treats the variables that contain factors or strings as 'id.vars', which remain fixed; and melts all numeric columns.
We start by making a mock table:

```{r }
subject <- c("Tim", "Ann", "Jill")
sex <- c("M", "F", "F")
control <- c(23, 31, 30)
experiment_1 <- c(34, 38, 36)
experiment_2 <- c(40, 42, 44)
df<-tibble(subject, sex, control, experiment_1, experiment_2)
df
```

Next we melt it by providing the df as the only argument to reshape2::melt
not that as an alternative to pipe ( %>% ) we can use
df_melted <- melt(df) %>% as_data_frame() #this gives identical result.
Wether to use piping or not is entirely up to you.

```{r}
df_melted <- df %>% melt() %>% as_tibble()
df_melted
```

Here we are more explicit about arguments to melt(). If you provide only id.vars or measure.vars, R will assume that all other variables belong to the argument that was not provided. 

```{r}
df_melted2 <- df %>% melt(id.vars=c("subject","sex"), # the variables not to split  
                          measure.vars=c("control", "experiment_1", "experiment_2"),
                          variable.name="experiment", # Name of the destination column for factors that are taken from names of melted columns
                          value.name="nr.of.counts" # name of the newly made column which contains the values
)
df_melted2
```


Alternatively we can use tidyr::gather to melt tables. Here you must provide three parameter values.
The 1st argument (here *key=experiment*) names the key factor or character column, whose values will be the names of the columns, which are melted into a single column. 
The 2nd argument (here *value=value*) is the name of the resultant single column, which contains the values. 
The third argument (here *3:ncol(df)*) specifies the columns that are melted into a single column; in the version *c(-subject, -sex)* every column except these 2 is melted. 

```{r}
df_melted3<-df %>% gather(key=experiment, value=value, 3:ncol(df))
# df_melted3<-df %>% gather(experiment, value, 3:ncol(df)) works as well.
df_melted4<-df %>% gather(experiment, value, c(-subject, -sex))
df_melted3
```


And now we cast our melted table back into the wide format.
Here *variable* is the factor or character column, whose values will be transformed into column names and *value* is the name of the column, which contains all the values that are spread into the new columns.

```{r}
df_w3 <- df_melted %>% spread(key=variable, value=value)
df_w3
```




### The 5 verbs of dplyr: select, filter, group_by, summarise, mutate

NB! Check the data wrangling cheatsheet and dplyr help for further details

#### select() columns

Select selects, renames, and re-orders columns

To select columns from sex to value:
```{r}
library(dplyr)
df_melted
df_melted %>% select(sex:value)
select(df_melted, -(sex:value)) #selects everything, except cols from sex to value

```


To select 3 columns and rename *subject* to *SUBJ*:
```{r}
df_melted %>% select(sex, value, SUBJ=subject)
```


To select all cols, except sex and value, and rename the *subject* col:
```{r}

df_melted %>% select(-sex, -value, SUBJ=subject)
```

**helper functions you can use within select():**

starts_with("abc"): matches names that begin with ???abc???.

ends_with("xyz"): matches names that end with ???xyz???.

contains("ijk"): matches names that contain ???ijk???.

matches("(.)\\1"): selects variables that match a regular expression. This one matches any variables that contain repeated characters. You???ll learn more about regular expressions in strings.

num_range("x", 1:3) matches x1, x2 and x3.

```{r}
iris<-as_tibble(iris)
select(iris, starts_with("Petal"))
select(iris, ends_with("Width"))

# Move Species variable to the front
select(iris, Species, everything())

df <- as.data.frame(matrix(runif(100), nrow = 10))
df <- tbl_df(df[c(3, 4, 7, 1, 9, 8, 5, 2, 6, 10)])
select(df, V9:V6)
select(df, num_range("V", 9:6))

# Drop variables with -
select(iris, -starts_with("Petal"))

# Renaming -----------------------------------------
# * select() keeps only the variables you specify
# * rename() keeps all variables
rename(iris, petal_length = Petal.Length)
```

See ?select for more details.


#### filter()

Filter filters rows


Keep rows that have sex level "M" and value >30.
```{r}
df_melted %>% filter(sex=="M" & value < 30)

```

Keep rows that have sex level "M" or value >30.
```{r}
df_melted %>% filter(sex=="M" | value < 30)
```


Keep rows that have sex level not "M" (which in this case equals "F") or value >30.
```{r}
df_melted %>% filter(sex != "M" | value <= 30)
```


Filtering with regular expression: we keep the rows where *subject* starts with the letter "T"

```{r}
library(stringr)
df_melted %>% filter(str_detect(subject, "^T")) 
#df_melted %>% filter(subject==(str_subset(subject, "^T"))) #same thing
```

As you can see there are endless vistas here, open for a regular expression fanatic. I so wish I was one!

remove NAs with filter()
```{r, eval=FALSE}
flights %>% filter(!is.na(dep_delay), !is.na(arr_delay))
```


#### summarise()

Summarise does just that - many rows summarised to a single value


```{r}
df_melted %>% summarise(MEAN= mean(value), SD= sd(value), MAD=mad(value), N= n(), unique_values_sex= n_distinct(sex))
```

To do something more exiting we must first group our observations by some factor(s) levels.

#### group_by()

Groups values for summarising or mutating

When we summarise by *sex* we will get two values for each summary statistic: for males and females. Aint that sexy?!

```{r}
df_melted %>% group_by(sex) %>% summarise(MEAN= mean(value, na.rm=TRUE), SD= sd(value), MAD=mad(value), N= n(), unique_values_sex= n_distinct(sex))
```


Now we group first by *variable* and then inside each group again by *sex*. This is getting complicated ...

```{r}
df_melted %>% group_by(variable, sex) %>% summarise(MEAN= mean(value), SD= sd(value), MAD=mad(value), N= n(), unique_values_sex= n_distinct(sex))
```

Now we group first by sex and then by variable. Spot the difference!

```{r}
df_melted %>% group_by(sex, variable) %>% summarise(MEAN= mean(value), SD= sd(value), MAD=mad(value), N= n(), unique_values_sex= n_distinct(sex))
```

*pro tip* if you want to summarise and then display the summary values as new column(s), which are added to the original non-shrunk df, use mutate() instead of summarize().

```{r}
df_melted %>% group_by(sex, variable) %>% mutate(MEAN= mean(value), SD= sd(value))
```

#### mutate()

However, the main use of the mutate() verb is for creating a new column for your table while taking information row-wise. The number of rows does not change while the number of columns increases. For example, if in your data frame called 'df' you have a column called 'value', you can create a new log2 transformed value value column called log_value by `df %>% mutate(log_value = log2(value))`. Or you can create a new column where a constant is substracted from the value column likel this: `df %>% mutate(centered_value = value - mean(value) ) `. Here the mean value is substracted from each individual value.


**Mutate adds new columns (and transmute() creates new columns while losing the previous columns - see the cheatsheet and help)**

Here we  firstly create a new column, which contains log-transformed values from the *value* column, and name it *log_value*.  
```{r}
df_melted %>% mutate(log_value = log10(value))
```

The same with transmute: note the dropping of some of the original cols, keeping the original *subject* col and renaming the *sex* col.

```{r}
df_melted %>% transmute(subject, gender=sex, log_value = log10(value))
```

```{r, eval=FALSE}

flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
) %>% 
mutate(
  gain = arr_delay - dep_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)

#you can refer to columns that you???ve just created:


```


*mutate_all(), mutate_if() and mutate_at() and the three variants of transmute() (transmute_all(), transmute_if(), transmute_at()) make it easy to apply a transformation to a selection of variables. See help.*


Here we first group and then mutate. Note that now, instead of a single constant, we divide by as many different constant as there are discrete factor levels in the sex variable (two, in our case)

```{r}
df_melted %>% group_by(sex) %>% mutate(normalised.value=value/mean(value), n2.val=value/sd(value))
```

Compare with a "straight" mutate to see the difference in values.

```{r}
df_melted %>% mutate(normalised.value=value/mean(value), n2.val=value/sd(value))
```

What are we doing here?

```{r, eval=FALSE}


not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    avg_delay1 = mean(arr_delay),
    avg_delay2 = mean(arr_delay[arr_delay > 0]) # the average positive delay
  )

# When do the first and last flights leave each day?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first = min(dep_time),
    last = max(dep_time)
  )

# Which destinations have the most carriers?
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers = n_distinct(carrier)) %>% 
  arrange(desc(carriers))

#Counts are so useful that dplyr provides a simple helper if all you want is a count:

not_cancelled %>% 
  count(dest)

#You can provide a weight. you could use this to ???count??? (sum) the total number of miles a plane flew:

not_cancelled %>% 
  count(tailnum, wt = distance)
```


Counts and proportions of logical values: sum(x > 10), mean(y == 0). When used with numeric functions, TRUE is converted to 1 and FALSE to 0. This makes sum() and mean() very useful: sum(x) gives the number of TRUEs in x, and mean(x) gives the proportion.

```{r, eval=FALSE}
# How many flights left before 5am? 
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(n_early = sum(dep_time < 500))

# What proportion of flights are delayed by more than an hour?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(hour_perc = mean(arr_delay > 60))
```


### Grouped filters

Find all groups bigger than a threshold:
```{r, eval=FALSE}
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)
```

If you need to remove grouping, and return to operations on ungrouped data, use ungroup().
```{r, eval=FALSE}
daily %>% 
  ungroup() %>%  # no longer grouped by date
  summarise(flights = n())
```


`str_replace_all()` helps to deal with unruly labelling inside columns containing strings

The idea is to find a pattern in a collection of strings and replace it with something else. 
String == character vector.

To find and replace we use str_replace_all(), whose base R analogue is gsub().

```{r}
library(stringr)
(bad.df<-tibble(time=c("t0", "t1", "t12"), value=c(2,4,9)))
(bad.df$time<-str_replace_all(bad.df$time, "t", "") %>% as.numeric()) 

```

now we have a numeric time column, which can be used in plotting.

or

```{r readr}

(bad.df$time<- parse_number(bad.df$time))

```

Here we did the same thing more elegantly by directly parsing numbers from a character string.

### gather()


### spread()


### separate() one column into several

```{r}
(df<-tibble(country=c("Albania"), disease.cases=c("80/1000")))
(df.sep<-df %>% separate(disease.cases, into=c("cases", "thousand")))
(df.sep<-df %>% separate(disease.cases, into=c("cases", "thousand"), sep="/"))
(df.sep<-df %>% separate(disease.cases, into=c("cases", "thousand"), sep=2))
(df.sep<-df %>% separate(disease.cases, into=c("cases", "thousand"), sep= -6))

(df<-tibble(index=c(1,2), taxon=c("Procaryota; Bacteria; Alpha-Proteobacteria; Escharichia", "Eukaryota; Chordata")))
(d1 <- df %>% separate(taxon, c('riik', 'hmk', "klass", "perekond"), sep = '; ', extra = "merge", fill="right")) 

# some special cases:
(df<-tibble(index=c(1,2), taxon=c("Prokaryota || Bacteria || Alpha-Proteobacteria || Escharichia", "Eukaryota || Chordata")))
(d1 <- df %>% separate(taxon, c("riik", "hmk", "klass", "perekond"), sep= "\\|\\|", extra = "merge", fill="right")) 

df <- tibble(index=c(1,2), taxon=c("Prokaryota.Bacteria.Alpha-Proteobacteria.Escharichia", "Eukaryota.Chordata"))
(d1 <- df %>% separate(taxon, c('riik', 'hmk', "klass", "perekond"), sep = '[.]', extra = "merge", fill="right")) 
#[] for points only.

(df<-tibble(index=c(1,2), taxon=c("Prokaryota.Bacteria,Alpha-Proteobacteria.Escharichia", "Eukaryota.Chordata")))
(d1 <- df %>% separate(taxon, c('riik', 'hmk', "klass", "perekond"), sep = '[,\\.]', extra = "merge", fill="right")) 
# [,\\.] separates by dot or comma. Isn't that cool?
```

The companion FUN to separate is `unite()` - see help.

