---
title: "Factors"
output: html_document
---

> factors are used to work with categorical variables, which have a fixed and known set of possible values. They are also useful when you want to display character vectors in a non-alphabetical order.

```{r}
library(tidyverse)
library(forcats)
```


Imagine that you have a variable that records month:
```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
```

Using a string to record this variable has two problems:

There is nothing saving you from typos:
```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
```

It doesn???t sort in a useful way:
```{r}
sort(x1)
```

To create a factor create a list of the valid levels:
```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```


Now you can create a factor:
```{r}
y1 <- factor(x1, levels = month_levels)
y1
```


```{r}
y2 <- factor(x2, levels = month_levels)
y2
```

alternatively
```{r}
y2 <- parse_factor(x2, levels = month_levels)
```

If you omit the levels, they???ll be taken from the data in alphabetical order

if you prefer the order of the levels matching the order of the first appearance in the data: 
```{r}
f2 <- x1 %>% factor() %>% fct_inorder()
f2
```

```{r}
levels(f2)
```

When factors are stored in a tibble, you can see levels with count():
```{r}
gss_cat
gss_cat %>%
  count(race)
levels(gss_cat$race)
```

##Modifying factor order

```{r}
gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  ) %>% 
ggplot(aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```

you could rewrite as:

```{r, eval=FALSE}
relig_summary %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig)) +
    geom_point()
```

####fct_relevel() pulls individual factor levels to the front. 

It takes a factor, f, and then any number of levels that you want to move to the front of the line.
```{r}
gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  ) %>% 
ggplot(aes(tvhours, fct_relevel(relig, "None", "Don't know"))) +
  geom_point()
```

####when you are colouring the lines on a plot, fct_reorder2() reorders the factor by the y values associated with the largest x values. 

This makes the plot easier to read because the line colours line up with the legend.

```{r}
gss_cat %>%
  filter(!is.na(age)) %>%
  group_by(age, marital) %>%
  mutate(N=n()) %>% 
ggplot(aes(age, N, colour = fct_reorder2(marital, age, N))) +
  geom_line() +
  labs(colour = "marital")
```

####for bar plots, you can use fct_infreq() to order levels in increasing frequency:
```{r}
gss_cat %>%
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(marital)) +
    geom_bar()
```

##Modifying factor levels

####fct_recode() allows to recode the value of each level.

```{r}
gss_cat %>% count(partyid)
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
  )) %>%
  count(partyid)
```

fct_recode() will leave levels that aren???t explicitly mentioned as is

you can assign multiple old levels to the same new level:
```{r, eval=FALSE}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party"
  )) %>%
  count(partyid)
```

####fct_collapse() is a useful variant of fct_recode(). 

For each new variable, you can provide a vector of old levels:
```{r, eval=FALSE}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

to lump together all the small groups to make a plot or table simpler.
the n parameter specifies how many groups (excluding other) we want to keep:
```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n = 5)) %>%
  count(relig, sort = TRUE) %>%
  print()
```


