---
title: "models"
output: html_notebook
---


world --> word (theory) --> equation (model)

1. every statistic is calculated within a model - there is no statistics that is independent of models

2. model is not the same thing as theory/hypothesis

if a theory directly corresopnds to reality, then this theory is true

a model cannot directly correspond to reality.

a model can directly correspond to a theory

model is an equation that tries to describe a theory (math that corresponds to words)



```{r}
x <- 1:100
y <- x
plot(y~x, type="l")
```

```{r}
x <- 1:100
a <- 30
y <- a+ x
plot(y~x, type="l", ylim=c(0, 100))
abline(c(0,1), lty=2)
```

```{r}
x <- 1:100
b <- 0.5
y <- b*x
plot(y~x, type="l", ylim=c(0, 100))
abline(c(0,1), lty=2)
```

```{r}
x <- 1:100
a <- 20
b <- 0.5
y <- a + b*x
plot(y~x, type="l", ylim=c(0, 100))
abline(c(0,1), lty=2)
```

```{r}
data(iris)
#head(iris)
m <- lm(data=iris, Sepal.Length~Petal.Length)
plot(iris$Sepal.Length~iris$Petal.Length, col=iris$Species)
abline(m)
```

```{r}
coef(m)
```

```{r}
Petal_length <- seq(min(iris$Petal.Length), max(iris$Petal.Length), length.out = 10)
a <- coef(m)[1]
b <- coef(m) [2]
Sepal_length <- a + b*Petal_length

plot(Sepal_length~Petal_length, type="b")
```

```{r}
m <- lm(data=iris, Sepal.Length~Sepal.Width)
m1 <- lm(data=iris, Sepal.Length~Sepal.Width -1)
coef(m); coef(m1)
AIC(m, m1)
```

```{r}
m2 <- lm(data=iris, Sepal.Length~Sepal.Width + Petal.Length)
coef(m); coef(m2)
```

```{r}
Sepal_width <- seq(min(iris$Sepal.Width), max(iris$Sepal.Width), length.out = 10)
Petal_length <- median(iris$Petal.Length)
Petal_length2 <- Petal_length + 1
a <- coef(m2)[1]
b1 <- coef(m2)[2]
b2 <- coef(m2)[3]

Sepal_length <- a + b1*Sepal_width + b2*Petal_length
Sepal_length2 <- a + b1*Sepal_width + b2*Petal_length2

plot(Sepal_length~Sepal_width, type="b")
lines(Sepal_length2~Sepal_width, lty=2)

```
```{r}
library(scatterplot3d)
iris1 <- iris[,1:3]
s3d <- scatterplot3d(iris1, angle=55, scale.y=0.7, pch=20,
                     col.axis = "blue", col.grid = "lightblue")
my.lm <- lm(data=iris, Sepal.Width~Sepal.Length + Petal.Length)
s3d$plane3d(my.lm, lty.box="solid")
```

